---
title: "Exploring the Wide World of ggplot2 Extensions"
author: 
  - "Eric R. Scott"
  - "Kristina Riemer"
  - "Renata Diaz"
date: 2024-06-20
format: 
  uaz-revealjs: default
execute: 
  echo: true
editor: visual
---

```{r}
library(ggplot2)
library(palmerpenguins)
```

## `gganimate`

Display transitions between states of a variable

Useful for showing trends in time series & spatial data

::: callout-tip
## Advice from the experts!

"Graphic elements should only transition between instances of the same underlying phenomenon"
:::

## Example

Body size distribution of penguins changing over the years

```{r}
ggplot(penguins, aes(x = body_mass_g, col = species)) +
  geom_density() +
  facet_wrap(~year)
```

## Show annual transition

```{r}
library(gganimate)

ggplot(penguins, aes(x = body_mass_g, col = species)) +
  geom_density() +
  transition_states(year) +
  ggtitle('Year: {closest_state}')
```

## Add more detail

```{r}
ggplot(penguins, aes(x = body_mass_g, col = species)) +
  geom_density() +
  transition_states(year) +
  ggtitle('Year: {closest_state}') +
  shadow_wake(wake = 0.25)
```

## Save file

```{r, eval=FALSE}
anim_save("~/Desktop/test_anim.gif")
```

## Resources

-   Cheat sheet: <https://rstudio.github.io/cheatsheets/gganimate.pdf>
-   Website: <https://gganimate.com>
